---
title: "final-project"
format: html
editor: visual
---

```{r}

library(tidyverse)
library(janitor)
library(stringr)
library(dplyr)
library(httr2)

census_key <- "2abeb09fab2a060893dafc5545972f25d26b0fb3"
url <- "https://api.census.gov/data/2021/pep/population"

request <- request(url) |>
  req_url_query(
    get = I("POP_2020,POP_2021,NAME"),
    `for` = I("state:*"),
    key = census_key
  )
```

```{r}
response <- request |> req_perform()
pop_2021 <- response |>
  resp_body_json(simplifyVector = TRUE) |>
  as_tibble()
pop_2021

str(pop_2021)
class(pop_2021)

```

```{r}

pop_2021_new <- pop_2021 |> row_to_names(row_number = 1)|>
  select(-state)|>
  # rename state column to state_name
  rename(state_name = NAME) |>
  # use pivot_longer to tidy
  pivot_longer(-state_name,
               names_to = "year",
               values_to = "population")|>
  # remove POP_ from year
  mutate(
    year = str_remove(year, "POP_"),
      # parese all relevant colunns to numeric
    year = as.numeric(year),
    population = as.numeric(population),
      # add state abbreviations using state.abb variable
      # use case_when to add abbreviations for DC and PR
    state = case_when(
      state_name == "District of Columbia" ~"DC",
      state_name == "Puerto Rico" ~"PR",
      TRUE ~ state.abb[match(state_name, state.name)]
    )
  )|> filter(year %in% c(2020, 2021)) |>
    arrange(state_name, population)
pop_2021_new
```

#Getting population data for 2022-23

```{r}
# Import the new population data
# https://www.census.gov/data/datasets/time-series/demo/popest/2020s-state-total.html 

population_new_raw <- read.csv("./data/raw/NST-EST2023-ALLDATA.csv")

# View the first few rows of the dataset
head(population_new_raw)

```

```{r}
#Wrangle the data

population_2223_clean <- population_new_raw |>
  filter(SUMLEV == 40) |>  # Keep only state-level data
  select("NAME", "POPESTIMATE2022", "POPESTIMATE2023") |>  # Select relevant columns
  rename(
    state_name = NAME,       # Rename NAME to state_name
    `2022` = POPESTIMATE2022, # Rename population columns for clarity
    `2023` = POPESTIMATE2023
  ) |>
  pivot_longer(
    cols = `2022`:`2023`,  # Convert population columns to long format
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(
    year = as.numeric(year), # Ensure year is numeric
    population = as.numeric(population) # Ensure population is numeric
  )

# Print cleaned dataset
print(population_2223_clean)

```

```{r}

full_population <- bind_rows(pop_2021_new, population_2223_clean) |>
  arrange(state_name, year)|>
  mutate(
    state = case_when(
      state_name == "District of Columbia" ~ "DC",
      state_name == "Puerto Rico" ~ "PR",
      is.na(state) ~ state.abb[match(state_name, state.name)],
      TRUE ~ state
    )
  )

# Print the combined dataset
print(full_population)

```

#Download covid case data

```{r}
api <- "https://data.cdc.gov/resource/pwn4-m3yp.json"
response <- request(api) |>
  req_url_query(`$limit` = 10000000000) |>  
  req_perform() 

cases_raw <- response |>
  resp_body_json()|>
  map_df(~ as_tibble(.))
```

```{r}
# wrangle covid case data
cases_clean <- cases_raw |>
  select(state, end_date, new_cases) |>
  rename(date = end_date, cases = new_cases) |>
  mutate(
    cases = as.numeric(cases),
    date = as_date(ymd_hms(date))
  ) |>
  mutate(mmwr_week = epiweek(date), mmwr_year = epiyear(date)) |>
select(state, mmwr_year, mmwr_week, cases) |>
arrange(state, mmwr_year, mmwr_week)

head(cases_clean)

```

# Get covid death and hospitalisation data

```{r}
get_cdc_data <- function(api){
  request(api) |>
  req_url_query("$limit" = 10000000) |>
  req_perform() |>
  resp_body_json(simplifyVector = TRUE)
}

hosp_raw <- get_cdc_data("https://data.cdc.gov/resource/39z2-9zu6.json")
deaths_raw <- get_cdc_data("https://data.cdc.gov/resource/r8kw-7aab.json")
vax_raw <- get_cdc_data("https://data.cdc.gov/resource/rh2h-3yt2.json")
```

# Wrangle the above data

```{r}
# Death
deaths <- deaths_raw |>
  filter(state %in% full_population$state_name) |>
  mutate(end_date = as_date(end_date),mmwr_year = epiyear(end_date)) |>
  rename(deaths_prov = covid_19_deaths,flu = influenza_deaths) |>
  mutate(mmwr_week = parse_number(mmwr_week),deaths = parse_number(deaths_prov)) |>
  filter(mmwr_year %in% c("2020", "2021","2022","2023", "2024"))|>
  select(state, mmwr_week, mmwr_year, deaths)

head(deaths)
```

```{r}
# hospitalisation 
hosp <- hosp_raw |>
filter(jurisdiction %in% full_population$state) |>
rename(hosp = new_covid_19_hospital, state = jurisdiction) |>
mutate(hosp = parse_number(hosp),
date = as_date(ymd_hms(collection_date)),
mmwr_week = epiweek(date), mmwr_year = epiyear(date)) |>
select(state, mmwr_year, mmwr_week, hosp) |>
group_by(state, mmwr_year, mmwr_week) |>
summarize(hosp = sum(hosp), n = n(), .groups = "drop") |>
filter(n == 7) |>
select(-n) |>
arrange(mmwr_year, mmwr_week)

head(hosp)
```

```{r}
# vaccination 
vax <- vax_raw |> filter(date_type == "Admin" & location %in% full_population$state) |>
rename(state = location, series_complete = series_complete_cumulative,
booster = booster_cumulative) |>
mutate(date = as_date(ymd_hms(date)),
mmwr_week = as.numeric(mmwr_week), mmwr_year = epiyear(date),
series_complete = parse_number(series_complete),
booster = parse_number(booster)) |>
select(state, date, mmwr_week, mmwr_year, series_complete, booster) |>
group_by(state, mmwr_week, mmwr_year) |>
summarize(series_complete = max(series_complete),
booster = max(booster),.groups = "drop") |>
arrange(state, mmwr_year, mmwr_week)

head(vax)

```

# Make dates data frame

```{r}
all_dates <- data.frame(date = seq(make_date(2020, 1, 25),
make_date(2024, 12, 31),
by = "week")) |>
mutate(date = ceiling_date(date, unit = "week", week_start = 7) - days(1))|>
mutate(mmwr_year = epiyear(date), mmwr_week = epiweek(date))
dates_and_pop <- cross_join(all_dates, data.frame(state =
unique(full_population$state))) |> left_join(full_population, by = c("state",
"mmwr_year" = "year"))

all_dates
```

# Combine the above data frame

```{r}
# get deaths dataset a state column
# Create a state mapping table
state_mapping <- tibble(
  state_name = state.name,
  state_abbr = state.abb
)

# Add entries for DC and Puerto Rico if not present
state_mapping <- state_mapping |> 
  add_row(state_name = "District of Columbia", state_abbr = "DC") |> 
  add_row(state_name = "Puerto Rico", state_abbr = "PR")

# Add state abbreviations to deaths dataset
deaths <- deaths |> 
  left_join(state_mapping, by = c("state" = "state_name"))

```

```{r}
# Use the updated deaths dataset with state abbreviations for joining
dat <- cases_clean |> 
  left_join(deaths, by = c("state"= "state_abbr", "mmwr_year", "mmwr_week")) |> 
  left_join(dates_and_pop, by = c("state", "mmwr_year", "mmwr_week")) |> 
  left_join(hosp, by = c("state", "mmwr_year", "mmwr_week")) |> 
  left_join(vax, by = c("state", "mmwr_year", "mmwr_week"))

head(dat)

```

## Q1 - Divide the pandemic period, January 2020 to December 2024 into waves. Justify your choice with data visualization.

```{r}
# Calculate rates and reshape the dataset
p <- dat |> 
  mutate(
    cases = cases / population * 100000,      # Calculate cases per 100,000
    hosp = hosp / population * 100000,        # Calculate hospitalizations per 100,000
    deaths = deaths / population * 100000     # Calculate deaths per 100,000
  ) |> 
  select(date, cases, hosp, deaths, state) |>  # Select relevant columns
  pivot_longer(
    cols = c(cases, deaths, hosp),            # Reshape the data
    names_to = "outcome", 
    values_to = "rate"
  ) |> 
  ggplot(aes(x = date, y = rate, color = state, group = state)) + 
  geom_line() + # Add line plot
  facet_wrap(~outcome, nrow = 3, scales = "free_y") + 
  labs(
    title = "COVID-19 Cases, Deaths, and Hospitalizations Over Time",
    x = "Date",
    y = "Rate per 100,000",
    color = "State"
  ) +
  theme_minimal()

# Print the plot
print(p)

```

# Segmentation of covid waves

```{r}
dat_wave <- dat |> 
  mutate(
    wave = case_when(
      # wave 1 is the initial outbreak globally 
      date >= as.Date("2020-01-01") & date < as.Date("2020-10-01") ~ "Wave 1",
      # wave 2 is the surge in fall and winter, where indoor gathering and holiday-related travel increased the cases; as well as the emergence of new variants like Alpha before vaccination became widely available.
      date >= as.Date("2020-10-02") & date < as.Date("2021-06-30") ~ "Wave 2",
      # wave 3 is when the contagious Delta variant began to circulate and eventually became dominant, hence significantly increased all metrics.
      date >= as.Date("2021-07-01") & date < as.Date("2021-11-30") ~ "Wave 3",
      # wave 4 is when Omicron BA.1 variant significantly increased cases number as it is highly transmissible but less severe; also due to the widespread vaccination.
      date >= as.Date("2021-12-01") & date < as.Date("2022-05-01") ~ "Wave 4",
      # wave 5 captures small wave associated with Omicron subvariants like BA.2, BA.5; less severe but more transmissible than Delta, also led to significant increases in hospitalisations and deaths.there are reduced activities across all metrics, potentially due to improved population immunity and pubic health measures. The peak highlights the virus’s ability to adapt and circulate under relaxed public health measures.
      date >= as.Date("2022-05-02")  ~ "Wave 5",
      TRUE ~ "Unknown"
    )
  )
```

```{r}
p_wave <- dat_wave |> 
  mutate(
    cases = cases / population * 100000,
    hosp = hosp / population * 100000,
    deaths = deaths / population * 100000
  ) |> 
  select(date, cases, hosp, deaths, state, wave) |> 
  pivot_longer(
    cols = c(cases, deaths, hosp), 
    names_to = "outcome", 
    values_to = "rate"
  ) |> 
  ggplot(aes(x = date, y = rate, color = wave, group = state)) + 
  geom_line() + 
  facet_wrap(~outcome, nrow = 3, scales = "free_y") +
  labs(
    title = "COVID-19 Waves by Outcome",
    x = "Date",
    y = "Rate per 100,000",
    color = "Wave"
  ) +
  theme_minimal()+
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 8)
  )

print(p_wave)

```

## Question 2 - For each period compute the deaths rates by state. Describe which states did better or worse during the different periods.

```{r}
# Summarise total deaths and calculate death rates
death_rates_by_wave <- dat_wave |>
  group_by(state, wave) |>
  filter(wave != "Unknown")|>
  summarize(
    total_deaths = sum(deaths, na.rm = TRUE),
    total_population = mean(population, na.rm = TRUE)
  ) |>
  mutate(death_rate = (total_deaths / total_population) * 100000) |>
  arrange(wave, desc(death_rate))

print(death_rates_by_wave)

```

```{r}
library(RColorBrewer)

# Visualise death rates across states and waves
ggplot(death_rates_by_wave, aes(x = wave, y = state, fill = death_rate)) +
  geom_tile(color = "white") +
  scale_fill_gradientn(    colors = brewer.pal(n = 9, name = "YlGnBu"), # Using a distinct Brewer palette
name = "Death Rate\n(per 100,000)") +
  labs(title = "Death Rates by State and Wave",
       x = "Wave",
       y = "State") +
  theme_minimal()+
  theme(
    axis.text.y = element_text(size = 6), # Adjust the font size for y-axis
    axis.text.x = element_text(size = 10), # Adjust font size for x-axis
    axis.title = element_text(size = 12) # Adjust font size for titles
  )
```

## Question 3 - Describe if COVID-19 became less or more virulent across the different periods.

```{r}
# 1. Case-Fatality Ratio (CFR) Over Time by Wave
# Step 1: Create a new dataset for Q3.1
dat_wave_q31 <- dat_wave |>
  # Filter out rows where cases <= 10 to avoid division by very small numbers
  filter(cases > 10) |>
  filter(wave != "Unknown")|>
  # Correct CFR calculation to avoid dividing by zero
  mutate(CFR = ifelse(cases > 0, (deaths / cases) * 100, NA)) 

# Step 2: Smooth the plot or limit the y-axis range to remove artifacts
p_cfr <- ggplot(dat_wave_q31, aes(x = date, y = CFR, color = wave, group = wave)) +
  geom_line(alpha = 0.6) + # Add transparency for better clarity
  scale_y_continuous(limits = c(0, 10)) + # Limit y-axis to a realistic range
  labs(title = "Case-Fatality Ratio (CFR) Over Time by Wave",
       y = "CFR (%)",
       x = "Date") +
  theme_minimal() +
  theme(legend.position = "right")

# Print the plot
print(p_cfr)
```

```{r}
# 2. Hospitalization-to-Death Ratio Over Time
# Step 1: Create a new dataset with HDR
dat_wave_hdr <- dat_wave |>
  filter(wave != "Unknown")|>
  mutate(HDR = ifelse(deaths > 0, hosp / deaths, NA)) |> # Calculate HDR
  group_by(date, wave) |>
  summarise(
    avg_HDR = mean(HDR, na.rm = TRUE),
    min_HDR = min(HDR, na.rm = TRUE),
    max_HDR = max(HDR, na.rm = TRUE)
  )

# Step 2: Generate the HDR plot over time
p_hdr <- ggplot(dat_wave_hdr, aes(x = date, y = avg_HDR, color = wave, group = wave)) +
  geom_line(size = 1) + # Add lines for each wave
  geom_ribbon(aes(ymin = min_HDR, ymax = max_HDR, fill = wave), alpha = 0.2, show.legend = FALSE) + # Add shaded area for range
  scale_y_continuous(trans = "log10", labels = scales::comma) + # Log scale for HDR
  labs(
    title = "Hospitalisation-to-Death Ratio (HDR) Over Time by Wave",
    x = "Date",
    y = "HDR (log scale)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
  )

# Print the plot
print(p_hdr)

```

```{r}
# 3. State-Level Comparison Heatmap
# Step 1: Create a dataset for the heatmap
state_wave_hdr <- dat_wave |>
  filter(wave != "Unknown")|>
  mutate(HDR = ifelse(deaths > 0, hosp / deaths, NA)) |> # Calculate HDR
  group_by(state, wave) |>
  summarise(avg_HDR = mean(HDR, na.rm = TRUE)) |> # Average HDR per state per wave
  ungroup()

# Step 2: Plot the heatmap
p_heatmap_hdr <- ggplot(state_wave_hdr, aes(x = wave, y = reorder(state, avg_HDR, FUN = median), fill = avg_HDR)) +
  geom_tile(color = "white", size = 0.2) + # Heatmap tiles with borders
  scale_fill_viridis_c(option = "C", name = "HDR (log scale)", trans = "log10", na.value = "grey80") + # Log scale for HDR
  labs(
    title = "State-Level Comparison of Hospitalisation-to-Death Ratio (HDR) by Wave",
    x = "Wave",
    y = "State"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "right", 
    axis.text.y = element_text(size = 6) # Adjust the font size for y-axis
  )

print(p_heatmap_hdr)

```

```{r}
# 4. Death Rates vs. Vaccination Coverage
# Step 1: Create a dataset for the scatter plot
death_vax <- dat_wave |>
  group_by(state, wave) |>
  filter(wave != "Unknown")|>
  summarise(
    death_rate = sum(deaths, na.rm = TRUE) / sum(population, na.rm = TRUE) * 100000, # Death rate per 100,000
    vax_rate = max(series_complete, na.rm = TRUE) / max(population, na.rm = TRUE) * 100 # Vaccination rate (percentage)
  ) |>
  ungroup()

# Step 2: Plot the scatter plot
p_death_vax <- ggplot(death_vax, aes(x = vax_rate, y = death_rate, color = wave, label = state)) +
  geom_point(size = 3, alpha = 0.8) + # Add points
  geom_smooth(method = "lm", se = FALSE, color = "black", linetype = "dashed") + # Add a linear regression line
  scale_color_viridis_d(name = "Wave") + # Discrete color scale for waves
  labs(
    title = "Death Rates vs. Vaccination Coverage by Wave",
    x = "Vaccination Coverage (%)",
    y = "Death Rate (per 100,000)"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12)
  )

# Add labels to points (optional, can be removed for cleaner plot)
p_death_vax <- p_death_vax +
  geom_text(size = 3, vjust = -1, hjust = 1, check_overlap = TRUE)

# Print the plot
print(p_death_vax)

```

```{r}
# 5. Rolling Averages of Deaths, Cases, and Hospitalisations
# Step 1: Compute rolling averages
dat_wave_roll <- dat_wave |>
  group_by(state) |>
  filter(wave != "Unknown")|>
  mutate(
    cases_avg = zoo::rollmean(cases, k = 7, fill = NA, align = "right"), # 7-day rolling average for cases
    deaths_avg = zoo::rollmean(deaths, k = 7, fill = NA, align = "right"), # 7-day rolling average for deaths
    hosp_avg = zoo::rollmean(hosp, k = 7, fill = NA, align = "right") # 7-day rolling average for hospitalizations
  ) |>
  ungroup()

# Step 2: Prepare dataset for visualization
dat_roll_long <- dat_wave_roll |>
  select(date, wave, cases_avg, deaths_avg, hosp_avg) |>
  pivot_longer(
    cols = c(cases_avg, deaths_avg, hosp_avg),
    names_to = "metric",
    values_to = "rolling_avg"
  )

# Step 3: Plot rolling averages
p_roll <- ggplot(dat_roll_long, aes(x = date, y = rolling_avg, color = wave)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ metric, scales = "free_y", nrow = 3, labeller = as_labeller(c(
    cases_avg = "Cases (7-day Avg)",
    deaths_avg = "Deaths (7-day Avg)",
    hosp_avg = "Hospitalisations (7-day Avg)"
  ))) +
  scale_color_viridis_d(name = "Wave") +
  labs(
    title = "Rolling Averages of Cases, Deaths, and Hospitalisations",
    x = "Date",
    y = "7-day Average"
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    strip.text = element_text(size = 12)
  )
print(p_roll)

```
